---
title: "Project_Mateusz_Kucharz"
author: "Mateusz Kucharz"
date: "2022-11-29"
output: html_document
---

## Data I will be working on
#### 1. Number of children per woman
#### 2. Income (GDP per capita)
#### 3. Gini coefficient 
#### 4. Human development index
#### 5. Expected lifespan
#### 6. Number of child deaths (per 1000 live births, within the first year of life) 
#### 7. Average number of years spent in school for individuals 25+ years (Men)
#### 8. Average number of years spent in school for individuals 25+ years (Women)

## Countries I will use during the project:
#### Norway, USA, Poland, Bulgaria, India, Tunisia, Senegal, Cameroon, Brazil, Guatemala
#### These have been chosen because they represent a mix of countries that are very highly developed, highly, moderately, and poorly developed

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(gapminder)
library(dplyr)
library(tidyr)
library(ggplot2)
attach(anscombe)

dzieci_na_kobiete<-read.csv("C:/Users/mkuch/Rfile/dzieci_na_kobiete.csv")
gdp<-read.csv("C:/Users/mkuch/Rfile/GDP.csv")
gini<-read.csv("C:/Users/mkuch/Rfile/gini.csv")
hdi<-read.csv("C:/Users/mkuch/Rfile/hdi.csv")
dl_zycia<-read.csv("C:/Users/mkuch/Rfile/oczek_dl_zycia.csv")
niemowle<-read.csv("C:/Users/mkuch/Rfile/noworodki.csv")
srednia_ile_lat_szkola_ch<-read.csv("C:/Users/mkuch/Rfile/srednia_ile_lat_w_szkole_ch.csv")
srednia_ile_lat_szkola_k<-read.csv("C:/Users/mkuch/Rfile/srednia_ile_lat_w_szkole_dz.csv")

dzieci_na_kobiete1<-gather(dzieci_na_kobiete, year, dzieci, -country)
gdp1<-gather(gdp, year, gdp, -country)
gini1<-gather(gini, year, ginni, -country)
hdi1<-gather(hdi, year, hdi, -country)
dlzycia1<-gather(dl_zycia, year, dlzycia, -country)
niemowle1<-gather(niemowle, year, zgonyDZ, -country)
sredniach1<-gather(srednia_ile_lat_szkola_ch, year, szkolaCH, -country)
sredniak1<-gather(srednia_ile_lat_szkola_k, year, szkolaK, -country)

dz_na_kob2<-separate(dzieci_na_kobiete1, col = year, 
                     into = c(NA,"year"), sep = "X")

gdp2<-separate(gdp1, col = year, 
               into = c(NA,"year"), sep = "X")

gini2<-separate(gini1, col = year, 
                into = c(NA,"year"), sep = "X")

hdi2<-separate(hdi1, col = year, 
               into = c(NA,"year"), sep = "X")

dl_zycia2<-separate(dlzycia1, col = year, 
                    into = c(NA,"year"), sep = "X")

niemowle2<-separate(niemowle1, col = year, 
                     into = c(NA,"year"), sep = "X")

srednia2ch<-separate(sredniach1, col = year, 
                     into = c(NA,"year"), sep = "X")

srednia2k<-separate(sredniak1, col = year, 
                    into = c(NA,"year"), sep = "X")


dane <-inner_join(dz_na_kob2,gdp2, by=c("country", "year"))

dane1 <-inner_join(dane,gini2, by=c("country", "year"))

dane2 <-left_join(dane1,hdi2, by=c("country", "year"))

dane3 <-inner_join(dane2,dl_zycia2, by=c("country", "year"))

dane4 <-inner_join(dane3,niemowle2, by=c("country", "year"))

dane5 <-inner_join(dane4,srednia2ch, by=c("country", "year"))

dane6 <-inner_join(dane5,srednia2k, by=c("country", "year"))

wybrane_kraje <-c("Norway", "United States", "Poland", "Bulgaria", "India", 
                  "Tunisia", "Senegal", "Cameroon", "Brazil", "Guatemala")

kraje <- dane6 %>%
  group_by(country) %>%
  mutate(year = as.numeric(year),dzieci = round(dzieci,0),ginni=round(ginni,0)/100)%>%
  filter(country %in% wybrane_kraje)%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))


projekt<-kraje%>%
  filter(year %in% seq(1972,2022, by=5))



#wykres1
dane_wykres1<-dane4%>% filter(year %in% seq(2000,2015, by=5))%>%
  group_by(country) %>%
  mutate(year = as.numeric(year),dzieci = round(dzieci,0),ginni=round(ginni,0)/100)%>%
  filter(country %in% wybrane_kraje)%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))

  
#wykres2
dane_wykres2<-dane4%>% 
  group_by(country) %>%
  mutate(year = as.numeric(year),ginni=round(ginni,0)/100)%>%
  filter(country %in% wybrane_kraje,year %in% seq(1962,2012, by=5))%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))%>%
  select(country,ginni, gdp)

dane_wykres2.1<-dane4%>% filter(year %in% seq(1962,2012, by=5))%>%
  group_by(country) %>%
  mutate(year = as.numeric(year))%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))%>%
  select(country,ginni, gdp)



x1<-dane_wykres2$ginni
y1<-dane_wykres2$gdp

srgi<-round(mean(x1),2)
srgd<-round(mean(y1),2)

srednia1 <- c(srgi,srgd)

sdgi<-round(sd(x1),2)
sdgd<-round(sd(y1),2)

odchylenie1 <- c(sdgi,sdgd)

cvgi<-round(sdgi/srgi * 100,2)
cvgd<-round(sdgd/srgd * 100,2)

zmiennosc1 <- c(cvgi, cvgd)

wykres2 <- data.frame(srednia1, odchylenie1, zmiennosc1)
rownames(wykres2)<-c("Gini","PKB")

#wykres3

dane_wykres3<-dane4%>% 
  group_by(country) %>%
  mutate(year = as.numeric(year))%>%
  filter(country %in% wybrane_kraje,year %in% seq(1962,2012, by=5))%>%
  select(country, year, gdp)%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))

Bulgaria<- dane_wykres3 %>% 
  filter(country == "Bulgaria")%>%select(country, gdp)
Brazylia<- dane_wykres3 %>% 
  filter(country == "Brazil")%>%select(country, gdp)
Kamerun<- dane_wykres3 %>% 
  filter(country == "Cameroon")%>%select(country, gdp)
Guatemala<- dane_wykres3 %>% 
  filter(country == "Guatemala")%>%select(country, gdp)
Indie<- dane_wykres3 %>% 
  filter(country == "India")%>%select(country, gdp)
Norewgia<- dane_wykres3 %>% 
  filter(country == "Norway")%>%select(country, gdp)
Polska<- dane_wykres3 %>% 
  filter(country == "Poland")%>%select(country, gdp)
Senegal<- dane_wykres3 %>% 
  filter(country == "Senegal")%>%select(country, gdp)
Tunezja<- dane_wykres3 %>% 
  filter(country == "Tunisia")%>%select(country, gdp)
USA<- dane_wykres3 %>% 
  filter(country == "United States")%>%select(country, gdp)

srednia2<-c(mean(Bulgaria$gdp),mean(Brazylia$gdp),mean(Kamerun$gdp),mean(Guatemala$gdp),mean(Indie$gdp),mean(Norewgia$gdp),
           mean(Polska$gdp),mean(Senegal$gdp),mean(Tunezja$gdp),mean(USA$gdp) )

odchylenie2<-c(sd(Bulgaria$gdp),sd(Brazylia$gdp),sd(Kamerun$gdp),sd(Guatemala$gdp),sd(Indie$gdp),sd(Norewgia$gdp),
           sd(Polska$gdp),sd(Senegal$gdp),sd(Tunezja$gdp),sd(USA$gdp) )

cv1<-sd(Bulgaria$gdp)/mean(Bulgaria$gdp) * 100
cv2<-sd(Brazylia$gdp)/mean(Brazylia$gdp) * 100
cv3<-sd(Kamerun$gdp)/mean(Kamerun$gdp) * 100
cv4<-sd(Guatemala$gdp)/mean(Guatemala$gdp) * 100
cv5<-sd(Indie$gdp)/mean(Indie$gdp) * 100
cv6<-sd(Norewgia$gdp)/mean(Norewgia$gdp) * 100
cv7<-sd(Polska$gdp)/mean(Polska$gdp) * 100
cv8<-sd(Senegal$gdp)/mean(Senegal$gdp) * 100
cv9<-sd(Tunezja$gdp)/mean(Tunezja$gdp) * 100
cv10<-sd(USA$gdp)/mean(USA$gdp) * 100
zmiennosc2<-c(cv1,cv2,cv3,cv4,cv5,cv6,cv7,cv8,cv9,cv10)



#wykres4
dane_wykres4<-dane4%>% filter(year %in% c(2000,1990, 1980,2010))%>%
  group_by(country) %>%
  mutate(year = as.numeric(year),dzieci = round(dzieci,0) )%>%
  filter(country %in% wybrane_kraje)%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))%>%
  select(country,year,dlzycia)

#wykres5

dane_wykres5<-dane4%>% 
  group_by(country) %>%
  mutate(year = as.numeric(year),hdi=round(hdi,2) )%>%
  filter(country %in% wybrane_kraje,year %in% seq(1994,2015, by=3))%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))%>%
  select(country,year,zgonyDZ, hdi)

#wykres6
dane_wykres6<-dane4%>% filter(year %in% c(1982,1997, 2012))%>%
  group_by(country) %>%
  mutate(year = as.numeric(year),dzieci = round(dzieci,0) )%>%
  filter(country %in% wybrane_kraje)%>%
  mutate(gdp = case_when(
    grepl("k",gdp) ~ as.numeric(substring(gdp,1,nchar(gdp)-1))*1000,
    TRUE ~ as.numeric(gdp)))%>%
  select(country,year,dzieci)

#wykres7

dane_wykres7<- kraje%>%
  select(country, year, szkolaCH, szkolaK)%>%
  filter(year %in% seq(1974,2009, by=5))


```

# Question 1: Is a high Human Development Index synonymous with a country's high GDP?
```{r}

dane_wykres1 %>%
  ggplot() +
  geom_point(mapping = aes(x = dane_wykres1$hdi, y = dane_wykres1$gdp, color = as.factor(dane_wykres1$country)), size = 4) +
  ggtitle("Is a high Human Development Index synonymous with a country's high GDP?") +
  labs(y = "PKB",
       x = "Human Development Index",
       color = "Kraje") + facet_grid(year~.)+
  ylim(0,65000)+theme_bw()


```
```{r}
cor(dane_wykres1$gdp,dane_wykres1$hdi, method = "spearman")
cor(dane_wykres1$gdp,dane_wykres1$hdi, method = "pearson")
```
From the chart, we can notice that over time the social development of the discussed countries steadily increased.
In some cases, even significantly over 15 years, taking into account the charts and the Pearson correlation coefficient
which in this case is ~0.99, it can be confidently stated that a large increase in HDI is also accompanied by a large increase in a country's GDP.

# Question 2: Does a country's high GDP equate to its low social differentiation?

```{r}
wykres2
```


```{r}
summary(dane_wykres2)
```


```{r}
round(cor(x1,y1,method = "spearman"),3)
round(cor(x1,y1,method = "pearson"),3)

```


```{r}
#Dla wszystkich państw z gapmindera
round(cor(dane_wykres2.1$ginni,dane_wykres2.1$gdp,method = "pearson"),3)
round(cor(dane_wykres2.1$ginni,dane_wykres2.1$gdp,method = "spearman"),3)
```

Because a high Gini index indicates a large social differentiation in a country, the low Pearson and Spearman correlation tells us that there is a very weak dependence between the Gini index and GDP. That is, if social differentiation deepens in a country, the GDP does not increase, in the case that we considered a larger number of countries, the dependencies should come out negative
this indicates that as the dependence decreases GDP should increase.

# Question 3: Checking the change in GDP size in selected countries over the years
```{r}
dane_wykres3 %>%
  ggplot(aes(x = year, y = gdp, col = country)) +
  geom_line(size = 1) + xlim(1962,2012)+
  ylim(0,65000)+theme_bw() 

```
```{r}
wyniki2 <- data.frame(srednia2, odchylenie2, zmiennosc2)
rownames(wyniki2)<-c("Bulgaria", "Brazylia", "Kamerun", "Gwatemala", "Indie", "Norwegia", "Polska", "Senegal", "Tunezja", "USA")
wyniki2
```
From the chart, one can easily determine which countries belong to which group of developing countries.
It is clear to see steadily and vigorously developing states like the USA and Norway, while the rest of the countries
are relatively close to each other. The GDP of Poland, Bulgaria, and Brazil in the 1990+ maintains an upward trend, while the countries
below them on the chart cannot boast the same. It is worth paying attention to the calculated descriptive statistics,
as you can notice that in most countries, the variability occurs at an average level, but there are quite large
standard deviations for very developed countries, and this results from their large increases over these 50 years.

# Question 4: How does the expected lifespan change in selected countries?

```{r}
dane_wykres4$kraj <- dane_wykres4$country %>%
  substr(start = 1, stop = 2) %>%
  toupper()

dane_wykres4 %>%
  ggplot(aes(y = reorder(kraj, dlzycia), 
             x = dlzycia),
         fill = continent) +
  geom_point(size = 3, col = "black") + 
  labs(y = "Kraje",x = "Life expectancy")+
  ggtitle("How does the expected lifespan change in selected countries?")+
  facet_grid(.~year)+theme_bw()
```

The expected lifespan usually grew every 10 years in each country. Of course, the biggest differences between years are owned by countries that are moderately and poorly developed, with Guatemala taking the lead here. Poland, the USA, and Norway already in 1980 showed a high expected lifespan, and since these are large values and we are talking about age, it is increasingly difficult to increase this value.


# Question 5: Checking the relationship between HDI and child mortality within the first year of life (per 1000 births)

```{r}
dane_wykres5 %>%
  ggplot() +
  geom_point(mapping = aes(x = hdi, y = zgonyDZ, color = as.factor(country)), size = 2) +
  ggtitle("Number of child deaths and its dependency on HDI (1994-2015, every 3 years)") +
  labs(y = "Child deaths within the first year of life (per 1000 births)",
       x = "Human Development Index",
       color = "Kraje")+theme_bw()
```

The lack of proper hygiene, doctors, or climate in this case makes a big difference.
Below an HDI of 0.5, we mostly have African countries where the number of child deaths starts to rise sharply.
There is a noticeable significant improvement in the quality of life and a reduction in the number of deaths in Tunisia and Brazil over these few
years. In the rest, despite increases in HDI, deaths remain at a similar level, meaning they are probably caused by pregnancy complications or accidents, not living conditions.

# Question 6: Is it true that over the years the birth rate decreases as parents prioritize their careers?

```{r}
dane_wykres6$kraj <- dane_wykres6$country %>%
  substr(start = 1, stop = 2) %>%
  toupper()

dane_wykres6 %>%
  ggplot() +
  geom_col(mapping = aes(x = reorder(kraj, -dzieci), y = dzieci), fill="light green")+
  ggtitle("Is it true that over the years the birth rate decreases as parents prioritize their careers?")+
  labs(y = "Children per woman",
       x = "Kraje")+
  facet_grid(.~year)+theme_bw()
```

As can be observed, the number of children born to women shows a downward trend or the numbers remain low. This may be due to the fact that women are becoming increasingly independent and aspire to have their own professional careers rather than just taking care of the home as was the case in the past. There is a commonly discussed demographic crisis not only in Poland but also worldwide. Of course, there are some exceptions such as some smaller African countries where a woman has a very large number of children, but due to a lack of proper hygiene and medical care, some unfortunately die.

# Question 7: Do people increasingly value education?
```{r}
wykres7 <-ggplot(data = dane_wykres7, aes(x=szkolaK, y=szkolaCH, color=country, label=year)) +
  geom_point() +
  xlab("Years in school women(25yo +)") +
  ylab("Years in school men(25 yo +)") +
  ggtitle("Do people increasingly value education?") +
  geom_abline(slope = 1, intercept = 0, color = "grey")+
  scale_x_continuous(breaks=seq(1,15,by=2))+
  scale_y_continuous(breaks=seq(1,15,by=2))+
  theme_light()

wykres7 + geom_text(size=3.5) 
```

A growing trend in education is observed in every country discussed, with the right side of the chart, in countries such as Poland, the USA, Bulgaria, and Norway, showing similar increases in education years between women and men. In contrast, in less developed countries, it's clear that men place more emphasis on education. Generally speaking, year after year, more people are placing greater importance on proper education. It's worth noting the increases on the women's side, which may directly relate to the previously discussed issue of lower birth rates.


